\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{fontawesome5}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

    \title{Developing a Parallelized Genetic Algorithm Library using OpenMP, MPI, and CUDA\\}

    \author{
        \IEEEauthorblockN{Solomiia Gadiichuk}
        \IEEEauthorblockA{\textit{Faculty of Applied Sciences} \\
        Ukrainian Catholic University\\
        Lviv, Ukraine \\
        hadiichuk.pn@ucu.edu.ua}
        \and
        \IEEEauthorblockN{Bohdan Zasymovych}
        \IEEEauthorblockA{\textit{Faculty of Applied Sciences} \\
        Ukrainian Catholic University\\
        Lviv, Ukraine\\
        zasymovych.pn@ucu.edu.ua}
        \and
        \IEEEauthorblockN{Marta Kosiv}
        \IEEEauthorblockA{\textit{Faculty of Applied Sciences} \\
        Ukrainian Catholic University\\
        Lviv, Ukraine\\
        kosiv.pn@ucu.edu.ua}
        \and
        \IEEEauthorblockN{Mykola Balyk}
        \IEEEauthorblockA{\textit{Faculty of Applied Sciences} \\
        Ukrainian Catholic University\\
        Lviv, Ukraine\\
        balyk.pn@ucu.edu.ua}
    }

    \maketitle

    \begin{abstract}
        While Genetic Algorithms (GAs) are powerful optimization tools, their high computational cost limits scalability. This paper details the development of a genetic algorithm library that includes shared-memory and accelerator-based parallelization strategies to solve this problem. We describe an implemented multi-threaded Generational GA, an implemented CUDA-accelerated GA, an implemented MPI-based Island Model, an implemented OpenMP Cellular GA, and an implemented OpenMP Differential Evolution algorithm. The performance of this library is evaluated using 3D mathematical surfaces and high-dimensional polygon image approximation tasks.
    \end{abstract}

    \begin{IEEEkeywords}
        Genetic Algorithms, Parallel Computing, Software Library, OpenMP, MPI, AVX, CUDA, Optimization
    \end{IEEEkeywords}

    \section{Introduction}
    Evolutionary algorithms are useful for solving complex optimization problems where traditional methods fail or take too much time. However, evaluating fitness functions for large populations over many generations requires a lot of computing power. 

    To solve this, different parallelization methods can be used, depending on the hardware and the specific algorithm. This project aims to develop a fast C++ library that includes four different algorithms: a classic Generational GA, an Island Model GA, a Cellular GA, and Differential Evolution (DE). Using tools like OpenMP, OpenMPI, oneAPI/oneTBB, and CUDA, we will adapt these algorithms to the best hardware options (shared memory, distributed memory, and GPUs). 

    This paper presents the current progress of the library. It details the implemented multi-threaded generational algorithm and CUDA backend, then outlines the theory and plans for future parallel models. The source code is publicly available at \faGithub\ \href{https://github.com/Martik-k/GALib-Parallel}{our GitHub}.

    \section{Background and Preliminaries}

    Genetic algorithms (GAs) are search and optimization methods inspired by natural evolution. Unlike standard algorithms that often get trapped in a local optimum, GAs use probability to explore large search spaces by evolving a population of possible solutions over several generations. This allows them to effectively escape local traps and find the true global maximum (or minimum) of a problem.

    \subsection{Representation and Fitness}
    Every possible solution in a GA is represented as a sequence of parameters called a \textit{genotype}. How this solution actually looks or works in the problem is the \textit{phenotype}. Each solution is evaluated using a \textit{fitness function}, $F(x)$, which gives a score based on how well it solves the problem. The main goal of the algorithm is to find a genotype with the highest (or lowest) fitness score.

    \subsection{Evolutionary Operators}
    Moving from one generation to the next uses three main steps \cite{b1}:
    \begin{itemize}
        \item \textbf{Selection:} The selection process determines which individuals from the current population will be chosen as parents to create the next generation.
        \item \textbf{Crossover:} Selected pairs exchange parts of their genome to create offspring. This combines the best traits from different parents.
        \item \textbf{Mutation:} It introduces random changes to the genetic material of the offspring solutions, helping to maintain diversity in the population and explore new regions of the search space.
    \end{itemize}

    \subsection{Schema Theory and Premature Convergence}
        The mathematical foundation of Genetic Algorithms is based on \textit{schemas}---templates that describe a subset of chromosomes sharing specific genes \cite{b2, b3}.

        \begin{itemize}
            \item \textbf{Binary Case:} A schema uses ones, zeros, and a "don't care" symbol ($*$). For example, the schema \texttt{1*0} matches both \texttt{110} and \texttt{100}.
            \item \textbf{Real-Valued Case:} In our implementation, because single-point crossover preserves exact numbers without blending them, floating-point values act as discrete symbols. Therefore, an exact numerical schema like $[1.52, *, 3.14]$ is valid and functions as a stable building block.
        \end{itemize}

        The growth of these schemas is described by Holland's Schema Theorem:
        \begin{equation}
            m(H, t+1) \ge m(H, t) \cdot \frac{f(H)}{\bar{f}} \left[ 1 - p_c \frac{\delta(H)}{L-1} - o(H)p_m \right]
        \end{equation}

        The components of this formula are defined as follows:
        \begin{itemize}
            \item $m(H, t)$: The number of instances of schema $H$ at generation $t$.
            \item $f(H)$: The average fitness of the schema $H$.
            \item $\bar{f}$: The average fitness of the entire population.
            \item $p_c$: The probability of crossover.
            \item $p_m$: The probability of mutation.
            \item $\delta(H)$: The defining length (the distance between the first and last fixed gene).
            \item $o(H)$: The order (the total number of fixed genes).
            \item $L$: The total length of the chromosome.
        \end{itemize}

    This theorem mathematically proves that schemas which are short (fixed genes are close together), low-order (few fixed genes), and possess above-average fitness will grow exponentially in the population over successive generations. By combining these highly fit building blocks, the algorithm efficiently avoids local traps and converges toward the global maximum.

    \section{Algorithmic Models}
    
    \subsection{Generational Genetic Algorithm}
    First, we implemented a basic Real-Coded Genetic Algorithm (RCGA). It uses continuous variables (double type) to find the minimum of mathematical functions. 

    The whole algorithm is written using C++ templates. This is very useful because it makes the library flexible and easy to reuse. We can easily apply different data types or new fitness functions without changing the main code.

    The algorithm uses the following main operations:
    \begin{itemize}
        \item \textbf{Initialization:} We create the first population randomly. We use local random number generators for each thread. This makes the code thread-safe and prevents race conditions.
        \item \textbf{Evaluation:} We calculate the fitness score for all individuals. This takes a lot of time, so we use OpenMP (\texttt{\#pragma omp parallel for}) to run it on multiple processor cores at the same time.
        \item \textbf{Selection:} We choose Tournament Selection. It is very fast because we do not need to sort the whole population. It is also easy to run in parallel.
        \item \textbf{Crossover:} We use Single-Point Crossover. It is simple and fast for the computer memory. It also keeps the exact good coordinates from the parents.
        \item \textbf{Mutation:} We use Gaussian Mutation. It adds a small random number to the genes. This helps the algorithm make small steps to find the exact minimum, and sometimes big jumps to escape bad local zones.
    \end{itemize}

    \subsection{Differential Evolution}
    Differential Evolution (DE) is implemented for efficient optimization of continuous variables. Unlike traditional mutation operators, DE generates new candidate solutions by taking the difference between randomly selected individuals in the population. The mutation strategy used is DE/rand/1, where for each individual $x_i$, a mutant vector $v_i$ is created as:
    \begin{equation}
        v_i = x_{r1} + F \cdot (x_{r2} - x_{r3})
    \end{equation}
    where $r1, r2, r3$ are distinct random indices, and $F$ is the differential weight (mutation scaling factor).

    After mutation, binomial crossover is applied between the mutant $v_i$ and the original $x_i$ to produce the trial vector $u_i$:
    \begin{equation}
        u_{i,j} = \begin{cases} 
        v_{i,j} & \text{if } rand_j \leq CR \text{ or } j = j_{rand} \\
        x_{i,j} & \text{otherwise}
        \end{cases}
    \end{equation}
    where $CR$ is the crossover rate, and $j_{rand}$ ensures at least one gene is taken from the mutant.

    Selection is performed by comparing the fitness of $u_i$ and $x_i$, keeping the better one. This adaptive search mechanism allows DE to converge faster than traditional GAs on many optimization problems. The implementation uses OpenMP for parallel evaluation of the population.

    \subsection{Island Model Genetic Algorithm}
    The Island Model, or coarse-grained parallel GA, partitions the total population into several semi-isolated subpopulations known as islands. Each island evolves independently according to its own local parameters, such as mutation and crossover rates. This independence allows different islands to explore diverse regions of the search space simultaneously, effectively acting as a mechanism to preserve genetic diversity and prevent premature convergence. 

    To facilitate global search, islands periodically exchange genetic material through a process called migration. This exchange is governed by a migration interval, defining the frequency of transfers, and a migration size, defining the number of individuals sent. During a migration event, a selection policy (such as elitism) identifies individuals to be sent to neighboring islands, while a replacement policy determines how incoming migrants are integrated into the local population, typically by replacing the least fit members. The connectivity between islands is defined by a migration topology, such as a ring or a fully connected graph, which determines the path and speed of information propagation across the system.

    \subsubsection{Parallelization and MPI Communication}
    The implementation utilizes a multi-tiered parallelization strategy to maximize hardware utilization across compute clusters. At the local level, shared-memory parallelism is employed to accelerate fitness evaluations and the creation of offspring within each island. At the distributed level, the Message Passing Interface (MPI) handles the coordination and data transfer between separate processes. 

    A key feature of the communication layer is its asynchronous design. By utilizing non-blocking MPI primitives and thread-safe circular buffers, islands can continue their evolutionary cycles without being stalled by network latency or waiting for synchronization with peers. Migrants are serialized into a portable binary format for efficient transmission and are processed by the destination island at the start of its next generation. Additionally, occasional global synchronization points allow for the identification and sharing of the absolute best solution found across the entire archipelago, ensuring that all nodes eventually converge on the global optimum.

    \subsection{Cellular Genetic Algorithm}
    In this model, the population is organized as a two-dimensional grid where each cell represents an individual solution. Instead of interacting with the entire population, individuals interact only with their local neighbors defined by the Von Neumann neighborhood (up, down, left, and right). This spatial structure helps maintain diversity in the population and reduces the risk of premature convergence.
    
    Since the population is arranged as a grid, different cells can be processed independently. This structure makes the algorithm suitable for parallel processing where multiple cells of the population can be evaluated and evolved simultaneously.

    \subsection{Differential Evolution with Hybrid OpenMP/AVX}
    We plan to implement Differential Evolution (DE) for faster optimization of continuous variables. Unlike basic mutation, DE creates new solutions by calculating the mathematical difference between existing individuals in the population. This helps the algorithm adapt its search direction and converge much faster.

    % \section{Benchmark Problems and Demonstrations}

    % \section{Experimental Setup}

    % \section{Results and Discussion}

    % \section{Conclusion}
    % This section will summarize the library's development progress, highlighting the capabilities of the implemented parallelization strategies across specific hardware constraints.

    \section*{Acknowledgment}

    \begin{thebibliography}{00}
        \bibitem{b1} ``Real-Coded Genetic Algorithm,'' \textit{Algorithm Afternoon}, [Online]. Available: \url{https://algorithmafternoon.com/genetic/real_coded_genetic_algorithm/}. [Accessed: Mar. 5, 2026].
        
        \bibitem{b2} O. Farenyuk, ``Genetic Algorithms'' unpublished lecture notes.

        \bibitem{b3} ``Holland's schema theorem,'' \textit{Wikipedia}, [Online]. Available: \url{https://en.wikipedia.org/wiki/Holland%27s_schema_theorem}. [Accessed: Mar. 5, 2026].
    \end{thebibliography}
\end{document}